Inference for Trans-dimensional Bayesian Models with Diffusive Nested Sampling

### 2014-07-27 Documents/Paper/ms.tex
New figure
Some edits

\begin{document}
\maketitle
\abstract{Many inference problems require us to infer the number $N$ of
objects in some region, along with their properties $\{\xx_i\}_{i=1}^N$,
from a dataset $\mathcal{D}$. These problems are typically solved using one
of the following two methods: i)
by executing a Monte Carlo algorithm (such as Nested Sampling) once for each
possible value of $N$,
and calculating the marginal likelihood or evidence as a function of $N$; or
ii) by doing a single run that allows the model dimension $N$ to change (such as
Markov Chain Monte Carlo with birth/death moves), and obtaining the posterior
for $N$ directly. In this paper we present a general approach to this problem
that uses trans-dimensional MCMC embedded {\it within} a Nested Sampling
algorithm, allowing us to explore the posterior distribution and calculate
the marginal likelihood (summed over $N$) even if the problem contains a phase
transition or other difficult features such as multimodality.
We present two example problems, finding sinusoidal signals in
noisy data, and finding and measuring galaxies in a noisy astronomical image.}

\section{Introduction}\label{sec:introduction}
\setlength{\parindent}{0cm}
\setlength{\parskip}{3mm}
Many inference problems have the following structure. There is some unknown
number, $N$, of objects in a region. Each of the $N$ objects has a property
$\xx$, which may be a single scalar value (for example, a mass), or a set of
values (e.g. a position and a mass).

We would like to assign a prior to the objects' properties $\{\xx_i\}_{i=1}^N$.
Since $N$ may be large, it is usually easier to assign an ``interim prior''
conditional on some hyperparameters $\hyperparams$, and then assign a prior to
$\hyperparams$. This kind of model is usually called {\it hierarchical}.

The prior for $N$, $\hyperparams$, and $\{\xx_i\}$ is usually factorised
in the following way:

@@

Here we have assumed the priors for $N$ and $\hyperparams$ are independent, and
the interim prior for $\{\xx_i\}$ is iid and does not depend on $N$.
Given data $\mathcal{D}$, we usually want to calculate properties of the
posterior distribution given by:
\begin{eqnarray}
p(N, \hyperparams, \{\xx_i\} | \mathcal{D}) \propto
p(N, \hyperparams, \{\xx_i\})
p(\mathcal{D} | N, \hyperparams, \{\xx_i\})\label{eq:bayes}
\end{eqnarray}
The general method used to calculate properties of the posterior distribution
is to generate samples from it using Markov Chain Monte Carlo (MCMC).

@@

In other studies, it has been common to do separate runs of Nested Sampling
$N$ fixed at various trial values. Then, the posterior for $N$ can be calculated
based on the estimates of the evidence obtained as a function of $N$. However,
in this paper, we use trans-dimensional MCMC moves to infer $N$. In other words,
$N$ is treated as just another parameter, and the marginal likelihood defined
in Equation~\ref{eq:evidence}, which we can compute, also includes the sum over
the unknown parameter $N$. Our motivation
for using Nested Sampling is not primarily to calculate $\mathcal{Z}$.
The motivation for using Nested Sampling, rather than just sampling the
posterior using trans-dimensional MCMC, is that the posterior often contains
difficult features, such as strong dependencies or multiple modes, that cause
problems with mixing. Diffusive Nested Sampling
(DNS) replaces the posterior distribution with an alternative target
distribution composed of a mixture of the prior distribution with other, more
constrained distributions. This is similar to the idea of simulated annealing,
where the target distribution is modified to something ``easier'', however,
the sequence of distributions used in Nested Sampling avoids some of the
problems with simulated annealing: in particular, there is no need to choose
a temperature schedule, and the method does not fail when phase transitions
are present.

@@

>>>>

\begin{document}
\maketitle
\abstract{Many inference problems involve inferring the number $N$ of
objects in some region, along with their properties $\{\xx_i\}_{i=1}^N$,
from a dataset $\mathcal{D}$. These problems are typically solved using one
of the following two methods: i)
by executing a Monte Carlo algorithm (such as Nested Sampling) once for each
possible value of $N$,
and calculating the marginal likelihood or evidence as a function of $N$; or
ii) by doing a single run that allows the model dimension $N$ to change (such as
Markov Chain Monte Carlo with birth/death moves), and obtaining the posterior
for $N$ directly. In this paper we present a general approach to this problem
that uses trans-dimensional MCMC embedded {\it within} a Nested Sampling
algorithm, allowing us to explore the posterior distribution and calculate
the marginal likelihood (summed over $N$) even if the problem contains a phase
transition or other difficult features such as multimodality.
We present two example problems, finding sinusoidal signals in
noisy data, and finding and measuring galaxies in a noisy astronomical image.}

\section{Introduction}\label{sec:introduction}
\setlength{\parindent}{0cm}
\setlength{\parskip}{3mm}
Many inference problems have the following structure. There is some unknown
number, $N$, of objects in a region. Each of the $N$ objects has a property
$\xx$, which may be a single scalar value (for example, a mass), or a set of
values (e.g. a two-dimensional position and a mass). These problems are often
challenging due to the unknown (and potentially large) dimensionality.

To do the inference, we must assign a prior to the objects' properties
$\{\xx_i\}_{i=1}^N$.
Since $N$ may be large, it is usually easier to assign an ``interim prior''
conditional on some hyperparameters $\hyperparams$, and then assign a prior to
$\hyperparams$. This kind of model is usually called {\it hierarchical}.
The prior for $N$, $\hyperparams$, and $\{\xx_i\}$ is usually factorised
in the following way:

@@

Here we have assumed the priors for $N$ and $\hyperparams$ are independent, and
the interim prior for $\{\xx_i\}$ is iid and does not depend on $N$.
Given data $\mathcal{D}$, we usually want to calculate properties of the
posterior distribution given by:
\begin{eqnarray}
p(N, \hyperparams, \{\xx_i\} | \mathcal{D}) \propto
p(N, \hyperparams, \{\xx_i\})
p(\mathcal{D} | N, \hyperparams, \{\xx_i\})\label{eq:bayes}
\end{eqnarray}
The general method used to calculate properties of the posterior distribution
is to generate samples from it using Markov Chain Monte Carlo (MCMC). The
structure of the model is depicted graphically in Figure~\ref{fig:pgm}.

@@

In other studies, it has been common to do separate runs of Nested Sampling,
with $N$ fixed at various trial values.
Then, the posterior for $N$ can be calculated
based on the estimates of the evidence obtained as a function of $N$. However,
in this paper, we use trans-dimensional MCMC moves to infer $N$. In other words,
$N$ is treated as just another parameter, and the marginal likelihood defined
in Equation~\ref{eq:evidence}, which we can compute, also includes the sum over
the unknown parameter $N$. Our motivation
for using Nested Sampling is not primarily to calculate $\mathcal{Z}$.
The motivation for using Nested Sampling, rather than just sampling the
posterior using trans-dimensional MCMC, is that the posterior often contains
difficult features, such as strong dependencies or multiple modes, that cause
problems with mixing. Diffusive Nested Sampling
(DNS) replaces the posterior distribution with an alternative target
distribution composed of a mixture of the prior distribution with other, more
constrained distributions, facilitating mixing between the multiple modes
and along degeneracy curves.
This is similar to the idea of simulated annealing,
where the target distribution is modified from the posterior to something
``easier'', however,
the sequence of distributions used in Nested Sampling avoids some of the
problems with simulated annealing: in particular, there is no need to choose
a temperature schedule, and the method does not fail when phase transitions
are present.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{sinewaves_likelihood.pdf}
\caption{The inference for $N$, the number of sinusoids, based on the
sinusoid data. The true value was $N=2$.
\label{fig:sinewaves_likelihood}}
\end{center}
\end{figure}
